Paper title

### 2021-02-22 Manuscript.Rmd
ms edits and data figures

In their foundational paper, @Hill1966-gv recognized that linked negatively
selected variants reciprocally impede the efficacy of population to remove
those mutations. In general, linked selection reduces the fixation probability
of advantageous mutations and increases that of deleterious mutations, compared
to expectations under single-locus dynamics [@Birky1988-jm]. Allele frequency
dynamics of linked selected variants are also predicted to deviate from
expectations without interference. Under a multiplicative fitness model, where
the fitness reduction of a haplotype carrying multiple deleterious variants is
equal to the product of the fitness reduction of each mutation independently,

One (!!) paragraph on epistasis, dominance, etc and overall importance. Thus,
there is a lot of interest in disentangling and distinguishing between these
processes.

@@

In addition to impacting surrounding neutral variation, selected variants can
interact and interfere with selection acting for or against other mutations. In
their foundational paper, @Hill1966-gv recognized that linked negatively
selected variants impede the efficacy of a population to remove those
mutations. Generally, linked selection reduces the fixation probability of
advantageous mutations and increases that of deleterious mutations, compared to
expectations under single-locus dynamics [@Birky1988-jm; @Barton1995-mj]. The
joint allele frequency dynamics of linked variants under selection are also
predicted to deviate from expectations without interference. Since haplotypes
that carry multiple deleterious mutations have lower fitness than haplotypes
that carry just one or the other under a multiplicative fitness model, we
should expect to see those mutation segregate on different haplotypes more
often than together, leading to negative, or repulsion, linkage disequilibrium
(LD), although the extent of LD depends non-trivially on the strength of
selection and probability of recombination separating loci. Nonetheless, this
provides a simple, testable hypothesis under the standard selection model of
additive effects within loci and multiplicative fitness between: we should
observe average signed LD to be negative between pairs of selected variants.

Aside from selective interference under a multiplicative fitness model, where
the relative fitness conferred by a haplotype carrying two deleterious
mutations is equal to the product of the fitnesses of haplotypes carrying
individual mutations, considerable attention has been given to
non-multiplicative deviations in the fitness function, known as epistasis.
Early work focused primarily on mutations of strong effect and explored models
of both synergistic and diminishing returns (or antagonistic or threshold)
epistasis under sexual and asexual reproduction, often in the limiting case of
free recombination between deleterious variants. Generally, for sexually
reproducing organisms, deleterious load is decreased under simple models of
synergistic epistasis compared to multiplicative fitness [@Kimura1966-gc;
@Kondrashov1995-va]. Pervasive epistasis has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], as well as driving incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero, with negative synergistic epistasis leading to an excess of negative LD
[@Kondrashov1982-sf; @Charlesworth1990-kw].

While epistasis is the most commonly considered nonlinear effect influencing
interactions between selected variants, non-additive effects *within* a
locus may also cause deviations from our baseline models, either with or
without epistasis. Dominance and multi-locus interactions are jointly important
in shaping the expected equilibrium allele frequencies and mutation load due to
strongly damaging disease mutations [@Clark1998-kq]. However, despite dominance
and partial recessiveness being appreciated as important factors shaping linked
variation [@Turelli2000-kz; @Zhao2016-bb], dominance has received less
attention than epistasis in multi-locus selection models, and its impact on the
joint segregation of negatively selected variants is poorly understood. There
is ample evidence that a large fraction of selected mutations are at least
partially recessive in humans [@xx], flies [@xx], and other species [@xx], and
that average levels of dominance vary with the strength of selection [@xx].

- Patterns of dominance
  - @Huber2018-cp, @Haldane1930-tt, @Kacser1981-nc

### @Ohta1969-ie and @Ohta1971-yd

- Diffusion models for the neutral two-locus decay of LD $\sigma_d^2$

### @Hallgrimsdottir2008-sq

-

### @Good2020-wj

-

- Neutrality, additive, dominance, epistasis, general diploid selection models
- Analytic results: Older stuff, like some basic results from HR, through more
  recent results in @Good2020-wj
  - recursions under neutrality [@Golding1984-pu; @Hudson2001-sg]
- Recent numerical approaches [@Ragsdale2017-gg; @Ragsdale2019-nt] and others that aren't me?
- Resorting to discrete forward simulations (a few examples)

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding regions, and
assess how well we can expect to discriminate between modes of selective
interactions from genome-wide patterns of LD.

@@

@@

@@

## Numerical solution

### Moment closure approach

### Implementation

- Numerical approach for solving the two-locus model with recombination,
  single-population demography, and general selection models, following
  @Jouganous2017-pq and @Ragsdale2019-nt.

### Validation

### Expected LD under inferred human demographic history

@@

**idea**: LD within genes partitioned by whether two mutations fall within
domains or are in different domains (distance as a proxy? or can we get
information about whether they lie in the same gene or not?)

@@

## Population size changes [[do what]]

- Effect of demography on these same statistics
  - Bottleneck with recovery
  - Human-like size histories (using @Speidel2019-nj)

## Data... main result in title

- Present data from loss-of-function (LOF), synonymous, and nonsynonymous
  variants from Thousand Genomes, both for pairs of loci within genes and for
  pairs of loci between nearby genes. Do we see any differences, and do the
  different patterns in African vs non-African populations tell us anything
  about the possible selection model for different categories of variants?

@@

(ref:fig3cap) \textbf{Dominance.} (A) $D^2$, and (B) $D$.

4. Effects of demography on LD (bottleneck and recovery vs expansion, following
   Tennessen model)

5. LOF, missense, and synonymous variants, within and between genes.

6. Missense variants in genes partitioned by conservation? Tolerance to
   nonsynonymous changes?

>>>>

In their foundational paper, @Hill1966-gv recognized that linked negatively
selected variants reciprocally impede the efficacy of population to remove
those mutations. In general, linked selection reduces the fixation probability
of advantageous mutations and increases that of deleterious mutations, compared
to expectations under single-locus dynamics [@Birky1988-jm]. Allele frequency
dynamics of linked selected variants are also predicted to deviate from
expectations without interference. Under a multiplicative fitness model, where
the fitness reduction of a haplotype carrying multiple deleterious variants is
equal to the product of the fitness reduction of each mutation independently,
we should expect to see those mutation segregate on different haplotypes more
often than together, leading to negative, or repulsion, linkage disequilibrium
(LD), although the extent of LD depends non-trivially on the strength of
selection and probability of recombination separating loci [@Hill1966-gv].

Non-additive effects further complicate our evolutionary models, including
dominance (i.e., interactions *within* loci) and epistasis (interactions
*between* loci), and both are thought to be widespread phenomena. A large
fraction of nonsynonymous coding mutations are thought to be at least partially
recessive [@Huber2018-cp; @Agrawal2011-wb], with average levels of dominance
correlating with strength of selection [@Kacser1981-nc]^[correct citation?],
and dominance plays an important role in shaping expected equilibrium allele
frequencies and mutation load of strongly damaging disease mutations
[@Clark1998-kq].^[maybe also cite @Haldane1930-tt] On the other hand, epistasis
differentially impacts deleterious load for asexually vs sexually reproducing
organisms [@Kimura1966-gc; @Kondrashov1995-va] and has been invoked as an
explanation for the evolutionary advantage of sex [@Kondrashov1982-sf;
@Charlesworth1990-kw; @Barton1998-lu], and epistatic interactions may also
drive incompatibilities that lead to postzygotic isolation during the process
of speciation [@Turelli2000-kz]. Within populations, epistasis is known to
cause signed LD to deviate dramatically from zero [@Kondrashov1995-va;
@Charlesworth1990-kw]. However, despite patterns of dominance being appreciated
as important factors shaping linked variation [@Turelli2000-kz; @Zhao2016-bb],
and the evolutionary importance of epistatic interactions, we currently lack
models for predicting correlation patterns between linked mutations under
general selection models.^[See @McVean2000-ox].

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding genes and domains,
and assess how well we can expect to discriminate between modes of selective
interactions using genome-wide patterns of LD.

@@

Many well-known properties of two-locus dynamics and equilibrium LD come from
early work that centered on the multi-locus diffusion approximation
[@Kimura1955-qe; @Hill1968-vu; @Ohta1969-ie; @Ohta1971-yd]. This includes the
result that genome-wide averages of signed LD are expected to be zero under
neutrality. Under a two-locus biallelic model, where the left locus allows
alleles $A$ and $a$ and the right locus has alleles $B$ and $b$, the standard
covariance measure of LD is defined as $D = f_{AB} - f_{A}f_{B}$ where $f_{AB}$
is the haplotype frequency of double-derived types carrying both $A$ and $B$,
and $f_{A}$ and $f_{B}$ are the marginal frequencies of the derived alleles at
each locus. This covariance decays due to both drift and recombination at a
rate proportional to the inverse of the effective population size and the
distance separating loci: \[\mathbb{E}[D]_{t+1} = \left(1 - \frac{1}{2N_e(t)} -
r \right)\mathbb{E}[D]_t.\] While $\mathbb{E}[D] = 0$, the variance of $D$ is
non-zero, and @Ohta1971-yd presented their famed result that the variance of
$D$ under neutrality and steady-state demography, normalized by the joint
heterozygosity of the two loci, is \[\sigma_d^2 =
\frac{\mathbb{E}[D^2]}{\mathbb{E}[f_A(1-f_A)f_B(1-f_B)]}\approx\frac{5 + \rho /
2}{11 + 13\rho / 2 + \rho^2 / 2},\] where $\rho = 4N_e r$.

Analytic progress beyond these well-known results has come haltingly. In the
1980s, recursions were developed to compute the two-locus sampling distribution
under neutrality, that is, the probability of observing given counts of
two-locus haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach
would continue to be developed and later form the foundation for the inference
of local recombination rates from population genetic data [@Hudson2001-sg;
@McVean2004-gj].

To include selection, however, there have not been many advances beyond the
Monte Carlo simulation approach taken by @Hill1966-gv, albeit with somewhat
more powerful computational resources and more sophisticated software for
performing flexible forward simulation (e.g., @Thornton2014-pn,
@Haller2019-vm). Analytic results for two-locus distributions under selection
are notoriously difficult, with a few notable flashes of progress.
@McVean2007-xd considered the effect of a recent sweep on patterns of LD for
loci nearby or spanning the locus under selection, and in an impressive recent
paper, @Good2020-wj presented analytic solutions for patterns of LD between
rare mutations under additive selection with epistasis. Nonetheless, such
approaches are confined to steady-state demography and constrained selection
models.

Numerical methods straddle the two approaches of expensive and noisy discrete
simulation and analytic solutions, providing a more efficient and practical
method to compute expectations of two-locus diversity measures under a wider
range of parameters and demographic scenarios.^[cite other numerical
approaches, and need to read: @Hallgrimsdottir2008-sq] @Ragsdale2017-gg solved
the two-locus diffusion equation with additive selection at either locus,
although that paper focused on the applicability of two-locus statistics to
demographic inference. More recently, @Ragsdale2019-nt extended the
@Hill1968-vu system for $\mathbb{E}[D^2]$ to compute arbitrary moments of the
distribution of $D$ for any number of populations connected by migration and
admixture. This paper also showed that such a moments-construction can be used
to solve for the two-locus sampling distribution within a single population,
though requires a moment-closure approximation for nonzero recombination and
selection. In the [Methods], I extend this same approach to model arbitrary
diploid selection, which encompasses dominance, epistasis, and other forms of
selective interactions between two loci.

@@

The two-locus sampling distribution is the direct analog to the single-locus
site frequency spectrum (SFS) of a given sample size. Instead of describing the
density or number of mutations with a given allele count out of $n$ samples,
the two-locus distribution $\Psi_n$ stores the density or number of pairs of
loci with observed haplotype counts, so that $\Psi_n(i, j, k)$ is the number
of pairs where we see $i$ $AB$ haplotypes, $j$ $Ab$ haplotypes, and $k$ $aB$
haplotypes. Here, we use upper case to denote the derived allele, so the number
of doubly ancestral haplotypes $ab$ is $n-i-j-k$.

- Diffusion equation for $\psi(x, y, z)$
- Sampling to get $\Psi_n$
- Point to @Ragsdale2019-nt supplement, with $\dot{\Psi}_n = \mathcal{D}\Psi_n \ldots$
- Note that the selection operator

@@

### Moment closure and implementation

@@

## Expected LD under inferred human demographic history

@@

**idea**: LD within genes partitioned by whether two mutations fall within
domains or are in different domains (distance as a proxy? or can we get
information about whether they lie in the same gene or not?)
- see @Yeang2007-gj, @Ivankov2014-tn, @Taverner2020-lk

@@

## The effect of population size changes on signed LD

- Effect of demography on these same statistics

  - Bottleneck with recovery and expansion
  - Human-like size histories (using @Speidel2019-nj)
  - For each (in own column): \@ref(fig:fig4)
    - Top panel: demography
    - Next panel: additive selection and synergistic epistasis
    - Next panel: dominance with h = 0, 0.2
    - Next panel: gene-based dominance and compensatory mutations

## Signed LD for mutations within and between genes

## Large positive signed LD for missense mutations within annotated protein
domains

- Present data from loss-of-function (LOF), synonymous, and nonsynonymous
  variants from Thousand Genomes, both for pairs of loci within genes and for
  pairs of loci between nearby genes \@ref(fig:fig5). Do we see any
  differences, and do the different patterns in African vs non-African
  populations tell us anything about the possible selection model for different
  categories of variants?

Some results:

- Within genes, both synonymous and missense mutations have LD slightly greater
  than zero, and are not significantly different from each other.
- On the other hand, LOF mutations have negative LD, although the small number
  and low frequency of LOF mutations in human populations means that the
  measurement is quite noisy.
- In fact, taking each population independently, all measurements have confidence
  intervals that overlap with zero
- But assuming independent measurements between populations (which is probably not
  correct, as populations have shared history and thus shared diversity patterns),
  can we say that synonymous and missense mutations have values significantly greater
  than zero and LOF mutations have values less than zero?

Within domains \@ref(fig:fig6):

- Missense and synonymous mutations have positive LD when looking at pairs
  within the same annotated domain.
- Between domains, both are reduced to be roughly zero
- Within domains, missense mutations have signed LD much larger than synonymous
  mutations

@@

(ref:fig3cap) **Dominance effects.** (A) $D^2$, and (B) $D$.

```{r fig4, echo=F, fig.cap="(ref:fig4cap)", fig.align="center"}
knitr::include_graphics("figures/fig4.pdf")
```

(ref:fig4cap) **The effects of demography on signed LD.**

```{r fig5, echo=F, fig.cap="(ref:fig5cap)", fig.align="center"}
knitr::include_graphics("figures/fig5.pdf")
```

(ref:fig5cap) **LD within and between protein-coding genes.**

```{r fig6, echo=F, fig.cap="(ref:fig6cap)", fig.align="center"}
knitr::include_graphics("figures/fig6.pdf")
```

(ref:fig6cap) **LD within and between annotated domains in protein-coding
genes.**

### 2021-02-22 SupportingInformation.Rmd
ms edits and data figures

>>>>

- Tables with data presented in figures 5 and 6
- Table with Tajima's D for syn, mis, lof mutations for all, within, or outside of domains

# Supplementary Tables

```{r tabS1, echo=F, tab.align="center"}
library(knitr)
library(kableExtra)
pops <- read.csv("tables/populations.txt", sep=";", header=TRUE)
knitr::kable(pops, "latex", booktabs=T,
    caption="Thousand Genomes Project population descriptions for populations used in this study.") %>%
kable_styling(latex_options = "HOLD_position")
```

```{r tabS2, echo=F, tab.align="center"}
library(knitr)
library(kableExtra)
taj <- read.csv("tables/tajimas_d.txt", sep=";", header=TRUE)
knitr::kable(taj, "latex", booktabs=T, longtable=T,
    linesep = c("", "", "", "", "", "", "", "", "\\addlinespace"),
    caption="Tamija's $D$ for classes of coding mutations, both within annotated domains and outside of domains.") %>%
kable_styling(latex_options = c("HOLD_position", "repeat_header"))
```

# Supplementary Figures

1) Accuracy of jackknife for varying sample size
2) Repeat of figures 2, 3, and 4 but for average $D$ instead of $\sigma_d^2$
3)
